---
title: "p8105_hw5_zlb2008"
author: "Zakari L. Billo"
date: "2025-11-09"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(p8105.datasets)

knitr::opts_chunk$set(
  fig.width = 8,
  fig.height = 7,
  fig.align = "center",
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

## Problem 1

Suppose you put 𝑛 people in a room, and want to know the probability that at least two people share a birthday. For simplicity, we’ll assume there are no leap years (i.e. there are only 365 days) and that birthdays are uniformly distributed over the year (which is actually not the case).

Write a function that, for a fixed group size, randomly draws “birthdays” for each person; checks whether there are duplicate birthdays in the group; and returns TRUE or FALSE based on the result.

```{r}
set.seed(1)

bday_match <- function(n) {
  
  birthdays <- sample(1:365, size = n, replace = TRUE)
  
  return(any(duplicated(birthdays)))
}
```

Next, run this function 10000 times for each group size between 2 and 50. For each group size, compute the probability that at least two people in the group will share a birthday by averaging across the 10000 simulation runs. Make a plot showing the probability as a function of group size, and comment on your results.

```{r}
n_sims <- 10000
group_sizes <- 2:50

probabilities <- sapply(group_sizes, function(n) {
  
    sim_results <- replicate(n_sims, bday_match(n))
    
    mean(sim_results)
    
})

results_df <- data.frame(
  n = group_sizes,
  probability = probabilities
  ) 

results_df |>
  ggplot(aes(x = n, y = probability)) +
  geom_line(color = "blue", size = 1) +
  geom_point(color = "blue") +
  labs(
    title = "Birthday Problem Simulation",
    x = "Group Size (n)",
    y = "Probability of at least one shared birthday"
  ) +
  scale_y_continuous(labels = scales::percent_format()) +
  theme_minimal() |>
  print()
```

The probability of a shared birhtday appears to be positive monotinic as a function of group size. At a group sizes 23 and 50 there is a 51.30% and 97.03% chance of at least one shared birthday, respectively. The growth is rapid because we aren't checking for a match with a specific birthday or birthdate. Instead, we are checking for any match between any two people in the group.

## Problem 2

